---
title: "Survival_Analysis_with_R.Rmd"
jupyter:
  jupytext:
  formats: ipynb,Rmd
text_representation:
  extension: .Rmd
format_name: rmarkdown
format_version: '1.2'
jupytext_version: 1.11.2
kernelspec:
  display_name: R
language: R
name: ir
---

```{r}
library(reshape2)
library(ggplot2)
library(survival)
```

```{r,eval=FALSE}
install.packages("SurvRegCensCov")
install.packages("eha")
install.packages("rms")
devtools::install_github("cran/smoothSurv")
# install.packages("smoothSurv")
install.packages("asaur")
```

# Discussion: Survival Analysis

### Simulated example

Let's simulate lightbulb life time in minutes using exponential distribution. 
Let's say we have A light bulb with rate 0.001 and B light bulb with rate 0.0015.
We will generate 100 bulbs each.

```{r}
 set.seed(1234)
 lightbulb<-function(n=100,hrate=0.001){rexp(n,hrate)}
 df=data.frame(A=lightbulb(100,0.001),B=lightbulb(100,0.0015))
```
 
There are many ways to look at this data. For example, how many survived 50 hours?
```{r}
 ggplot(melt(df<3000))+geom_bar()+aes(x=factor(Var2,levels=c("B","A")), fill=value)+xlab("")+coord_flip()

```

Or the counts of how many died in every 10 hour window.
```{r}
 dfdisc=apply(df,2,function(x)cut(x, c(seq(0,10000,by=600),10000)))
 mdfdisc=melt(dfdisc)
 mdfdisc$value = factor(mdfdisc$value ,labels=1:9,levels=c("(0,600]", "(600,1.2e+03]" ,"(1.2e+03,1.8e+03]", "(1.8e+03,2.4e+03]","(2.4e+03,3e+03]"  ,  "(3e+03,3.6e+03]"  ,"(3.6e+03,4.2e+03]","(4.2e+03,4.8e+03]","(4.8e+03,5.4e+03]" ))
ggplot(mdfdisc)+geom_bar()+aes(x=value, )+xlab("")+facet_grid(Var2~.)
```

Or, we can look at the hours directly.

```{r}
ggplot(melt(df))+geom_histogram(bins=30)+aes(x=value )+xlab("")+facet_grid(variable~.)
````

Depending on the research question the way to analyze the data will be different.

- For the binary problem, you can use logistic regression.
- For the counts you can use Poisson regression with offset.
- For the hours, it depends on the shape of the distribution.

First take away message: It is important is to understand that not all survival analysis requires a complex survival analysis model.


### NCCTG Lung Cancer Data

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.

- `inst`:	Institution code
- `time`:	Survival time in days
- `status`:	censoring status 
  - 1=censored, 
  - 2=dead
- `age`:	Age in years
- `sex`:	
  - Male=1 
  - Female=2
- `ph.ecog`:	ECOG performance score as rated by the physician. 
  - 0=asymptomatic, 
  - 1= symptomatic but completely ambulatory, 
  - 2= in bed <50% of the day, 
  - 3= in bed > 50% of the day but not bedbound, 
  - 4 = bedbound
- `ph.karno`:	Karnofsky performance score (bad=0-good=100) rated by physician
- `pat.karno`:	Karnofsky performance score as rated by patient
- `meal.cal`:	Calories consumed at meals
- `wt.loss`:	Weight loss in last six months (pounds)

This is a famous data of survival in a randomized trial comparing two treatments for ovarian cancer.

```{r}
library(survival)
library("survminer")
data(cancer, package="survival")
?cancer
```

```{r}
fit.surv <- survfit(Surv(time, status) ~ 1,data=cancer)
#plot(fit.surv, xlab = "Time",
#     ylab = "Estimated Probability of Survival")
ggsurvplot(fit.surv, data = cancer)
```

We can create the curve for different predictor variables.
For example, the data has a variable

- `sex`:	
  - Male=1 
  - Female=2
  
When we plot the curve separated by gender. The patients that do not have the residual seem to have a higher survival rate.

```{r}
fit.surv <- survfit(Surv(time, status) ~ sex,data=cancer)
#plot(fit.surv, xlab = "Time",
#     ylab = "Estimated Probability of Survival")
ggsurvplot(fit.surv, data = cancer)
```
You can perform a log-rank test to compare the survival of patients with and without residual disease, using the `survdiff()` function.

```{r }
logrank.test <- survdiff(Surv(time, status) ~ sex,data=cancer)
logrank.test
```

```{r }
fit.cox <- coxph(Surv(time, status) ~ sex,data=cancer)
summary(fit.cox)
```



### Weibull regression on NCCTG Lung Cancer Data

Survival in patients with advanced lung cancer from the North Central Cancer Treatment Group. Performance scores rate how well the patient can perform usual daily activities.

- `inst`:	Institution code
- `time`:	Survival time in days
- `status`:	censoring status 
  - 1=censored, 
  - 2=dead
- `age`:	Age in years
- `sex`:	
  - Male=1 
  - Female=2
- `ph.ecog`:	ECOG performance score as rated by the physician. 
  - 0=asymptomatic, 
  - 1= symptomatic but completely ambulatory, 
  - 2= in bed <50% of the day, 
  - 3= in bed > 50% of the day but not bedbound, 
  - 4 = bedbound
- `ph.karno`:	Karnofsky performance score (bad=0-good=100) rated by physician
- `pat.karno`:	Karnofsky performance score as rated by patient
- `meal.cal`:	Calories consumed at meals
- `wt.loss`:	Weight loss in last six months (pounds)

```{r}
?lung
```

K-M plot
```{r}
# https://bioconnector.github.io/workshops/r-survival.html
sfit <- survfit(Surv(time, status)~sex, data=lung)
survminer::ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
           legend.labs=c("Male", "Female"), legend.title="Sex",  
           palette=c("dodgerblue2", "orchid2"), 
           title="Kaplan-Meier Curve for Lung Cancer Survival", 
           risk.table.height=.15)

summary(sfit, times=c(100, 200, 500))
```

Checking for proportional hazard assumption.
plotting log time vs cloglog survival.  If the proportional hazard assumption hold the lines should be parallel.
```{r}
kmsubset<- function(data,subset){
  sfit.0 <- survfit(Surv(time, status)~1,subset=subset, data=data)
  time.0 <- sfit.0$time
  surv.0 <- sfit.0$surv
  cloglog.0 <- log(-log(surv.0))
  logtime.0 <- log(time.0)
  aa=list(cloglog.0,logtime.0)
  names(aa)<-c("cloglog","logtime")
  return(aa)
}
a=kmsubset(subset=lung$sex==1, data=lung)
b=kmsubset(subset=lung$sex==2, data=lung)
plot(a$cloglog ~ a$logtime, type="s", col="blue", lwd=2)
lines(b$cloglog ~ b$logtime, col="red", lwd=2, type="s")
#lines(range(tt), range(tt)*lam, col=pal(2)[2], lwd=3)
```

We can start with an exponential regression.
```{r}
exp.lung<-survreg(Surv(time, status)~ph.ecog+sex+age,lung, dist='exponential')
summary(exp.lung)
plot(sfit, conf.int=FALSE, lwd=3, col='gray', mark.time=FALSE, xmax=4000,
     xaxt='n', bty='n', las=1, xlab='Time', ylab="Survival")
lab <- seq(0,10,2)
axis(1, at=lab*365.25, lab=lab)
 lines(predict(exp.lung, newdata=list(sex=1,ph.ecog=1,age=60),type="quantile",p=seq(.01,.99,by=.01)),seq(.99,.01,by=-.01),col="blue")
  lines(predict(exp.lung, newdata=list(sex=2,ph.ecog=1,age=60),type="quantile",p=seq(.01,.99,by=.01)),seq(.99,.01,by=-.01),col="red")
```

Following example from Zhang (2016) (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5233524/pdf/atm-04-24-484.pdf  ) we will fit a Weibull regression.

```{r}
wei.lung<-survreg(Surv(time, status)~ph.ecog+sex+age,lung, dist='weibull')
summary(wei.lung)
```

Compared to the exponential fit the Weibull fit seems to be much better.
```{r}
plot(sfit, conf.int=FALSE, lwd=3, col='gray', mark.time=FALSE, xmax=4000,
     xaxt='n', bty='n', las=1, xlab='Time', ylab="Survival")
lab <- seq(0,10,2)
axis(1, at=lab*365.25, lab=lab)
 lines(predict(wei.lung, newdata=list(sex=1,ph.ecog=1,age=60),type="quantile",p=seq(.01,.99,by=.01)),seq(.99,.01,by=-.01),col="blue")
  lines(predict(wei.lung, newdata=list(sex=2,ph.ecog=1,age=60),type="quantile",p=seq(.01,.99,by=.01)),seq(.99,.01,by=-.01),col="red")
```

Unfortunately the estimated coefficients are not clinically meaningful.

```{r}

library(SurvRegCensCov)
ConvertWeibull(wei.lung,conf.level = 0.95)
wei.lung.alt<-WeibullReg(Surv(time,status)~ph.ecog+sex+age,data=lung,conf.level=0.95)
```

Weibull model with categorical variables can be checked for its adequacy by stratified Kaplan-Meier curves. A plot of log survival time versus log[–log(KM)] will show linear and parallel lines if the model is adequate

Weibull regression diagnostic plot.
If the model fits well, a plot of log survival time versus log[–log(KM)] will show linear and parallel lines.
```{r}
WeibullDiag(Surv(time,status)~sex,data=lung)
```

Another package for Weibull regression.
```{r}

library(eha)
lung.alt<-weibreg( Surv(time, status)~age+sex+ph.ecog, data=lung)
lung.alt
```

Plot of Hazard, cumulative hazard, density and survival functions for 80 years old.

```{r}
par(mfrow=c(2,2))
plot(lung.alt, fn=c("haz"),new.data=c(80,2,3))
plot(lung.alt, fn=c("cum"),new.data=c(80,2,3))
plot(lung.alt, fn=c("den"),new.data=c(80,2,3))
plot(lung.alt, fn=c("sur"),new.data=c(80,2,3))
```

Plot of survival functions for 80,60,40,20 years old.

```{r}
par(mfrow=c(2,2))
plot(lung.alt, fn=c("sur"),new.data=c(80,2,3))
plot(lung.alt, fn=c("sur"),new.data=c(60,2,3))
plot(lung.alt, fn=c("sur"),new.data=c(40,2,3))
plot(lung.alt, fn=c("sur"),new.data=c(20,2,3))
```

Goodness of fit test comparing the semi-parametric model to the parametric model
```{r}
phreg.lung<-phreg(Surv(time, status)~ph.ecog+sex+age,lung, dist='weibull')
coxreg.lung<-coxreg(Surv(time, status)~ph.ecog+sex+age,lung)
check.dist(coxreg.lung,phreg.lung)
```

ANOVA
```{r}

library(rms)
psm.lung<-psm(Surv(time, status)~ph.ecog+sex*age+ph.karno+pat.karno+meal.cal+wt.loss,lung, dist='weibull')
anova(psm.lung)
```

Visualization functions:
```{r, message=FALSE}
plot(anova(psm.lung),margin=c("chisq", "d.f.", "P"))
fastbw(psm.lung,rule="aic")

psm.lung1<-psm(Surv(time, status)~ph.ecog+sex*age,lung,dist='weibull')
ggplot(Predict(psm.lung1, age=seq(20,80,by=5),ph.ecog=c(0,1,2,3),sex=c(1,2)))


library(smoothSurv)
smooth.lung <- smoothSurvReg(Surv(time, status)~ph.ecog+sex*age,data=lung, init.dist='weibull')
cov<-matrix(c(0,1,2,3,1,2,2,2,20,30,40,70,20,60,80,140),ncol=4,byrow=FALSE)
cov

par(mfrow=c(2,2))
survfit(smooth.lung,cov=cov)
survfit(smooth.lung, cdf = TRUE,cov=cov)
hazard(smooth.lung,cov=cov)
fdensity(smooth.lung,cov=cov)
```



## Problem set 

### Brain Tumor
This exercise focuses on the brain tumor data, which is included in the ISLR2 R library.
```{r}
library(ISLR2)
attach(BrainCancer)
```

(a) Plot the Kaplan-Meier survival curve with ±1 standard error
bands, using the survfit() function in the survival package.

```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

(b) Draw a bootstrap sample of size n = 88 from the pairs (yi, δi),
and compute the resulting Kaplan-Meier survival curve. Repeat
this process B = 200 times. Use the results to obtain an estimate
of the standard error of the Kaplan-Meier survival curve at each
timepoint. Compare this to the standard errors obtained in (a).

```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

(c) Fit a Cox proportional hazards model that uses all of the predictors
to predict survival. Summarize the main findings.


```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

(d) Stratify the data by the value of ki. (Since only one observation
has ki=40, you can group that observation together with the observations
that have ki=60.) Plot Kaplan-Meier survival curves
for each of the five strata, adjusted for the other predictors.

```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

### Censoring

This example makes use of the data in 

Observation (Y ) |Censoring Indicator ($\delta$) | Covariate (X)
-----------------|-------------------------------|-------------------
26.5             | 1                             | 0.1
37.2             | 1                             | 11
57.3             | 1                             | -0.3
90.8             | 0                             | 2.8
20.2             | 0                             | 1.8
89.8             | 0                             | 0.4

(a) Create two groups of observations. In Group 1, X <2, whereas
in Group 2, X ≥ 2. Plot the Kaplan-Meier survival curves corresponding
to the two groups. Be sure to label the curves so that
it is clear which curve corresponds to which group. By eye, does
there appear to be a difference between the two groups’ survival
curves?

```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

(b) Fit Cox’s proportional hazards model, using the group indicator
as a covariate. What is the estimated coefficient? Write a sentence
providing the interpretation of this coefficient, in terms
of the hazard or the instantaneous probability of the event. Is
there evidence that the true coefficient value is non-zero?

```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

(c) Recall from Section 11.5.2 that in the case of a single binary
covariate, the log-rank test statistic should be identical to the
score statistic for the Cox model. Conduct a log-rank test to determine
whether there is a difference between the survival curves
for the two groups. How does the p-value for the log-rank test
statistic compare to the p-value for the score statistic for the
Cox model from (b)?

```{r,echo=TRUE}
#
#
```


Your answer:

~~~
Please write your answer in full sentences.


~~~

## Additional Material


### Distributions

```{r,fig.width=16, fig.height=4}
const=function(x,lambda){return(rep(lambda,length(x)))}
par(mfrow=c(1,4))
curve(dexp(x,2),from=0, to=5,xlab="T",ylab="exp(lambda=2)",main="density")
curve(1-pexp(x,2),from=0, to=5,xlab="T",ylab="exp(lambda=2)",main="CDF")
curve(pexp(x,2),from=0, to=5,xlab="T",ylab="exp(lambda=2)",main="Survival")
curve( const(x,2),from=0, to=5,xlab="T",ylab="exp(lambda=2)",main="hazard")

par(mfrow=c(1,4))
curve(dweibull(x,2,1),from=0, to=5,xlab="T",ylab="Weibull(shape=2,scale=1)",main="density")
curve(1-pweibull(x,2,1),from=0, to=5,xlab="T",ylab="Weibull(shape=2,scale=1)",main="CDF")
curve(pweibull(x,2,1),from=0, to=5,xlab="T",ylab="Weibull(shape=2,scale=1)",main="Survival")
curve(dweibull(x,2,1)/pweibull(x,2,1),from=0, to=5,xlab="T",ylab="Weibull(shape=2,scale=1)",main="hazard")

par(mfrow=c(1,4))
curve(dgamma(x,3,2),from=0, to=5,xlab="T",ylab="gamma(shape=3,rate=2)",main="density")
curve(1-pgamma(x,3,2),from=0, to=5,xlab="T",ylab="gamma(shape=3,rate=2)",main="CDF")
curve(pgamma(x,3,2),from=0, to=5,xlab="T",ylab="gamma(shape=3,rate=2)",main="Survival")
curve(dgamma(x,3,2)/pgamma(x,3,2),from=0, to=5,xlab="T",ylab="gamma(shape=3,rate=2)",main="hazard")

```